---
title: "Cyclistic Bike-share"
author: "Ng Tri"
date: '2022-03-14'
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

# # # # # # # # # # # # # # # # # # # # # # # 
# Install required packages
# tidyverse for data import and wrangling
# lubridate for date functions
# ggplot for visualization
# # # # # # # # # # # # # # # # # # # # # # #  

```{r}
library(tidyverse)
library(lubridate) 
library(ggplot2)  
library(dplyr)
setwd("D:\\DH\\3rd - Semester II\\Google Data Analytics\\Case study\\Data\\CSV")
```

#=====================
# STEP 1: COLLECT DATA
#=====================

```{r}
m1 <- read_csv("202101-divvy-tripdata.csv")
m2 <- read.csv("202102-divvy-tripdata.csv")
m3 <- read.csv("202103-divvy-tripdata.csv")
m4 <- read.csv("202104-divvy-tripdata.csv")
m5 <- read.csv("202105-divvy-tripdata.csv")
m6 <- read.csv("202106-divvy-tripdata.csv")
m7 <- read.csv("202107-divvy-tripdata.csv")
m8 <- read.csv("202108-divvy-tripdata.csv")
m9 <- read.csv("202109-divvy-tripdata.csv")
m10 <- read.csv("202110-divvy-tripdata.csv")
m11 <- read.csv("202111-divvy-tripdata.csv")
m12 <- read.csv("202112-divvy-tripdata.csv")
```

#====================================================
# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
#====================================================

# Compare column names each of the files

```{r}
colnames(m1)
colnames(m2)
colnames(m3)
colnames(m4)
colnames(m5)
colnames(m6)
colnames(m7)
colnames(m8)
colnames(m9)
colnames(m10)
colnames(m11)
colnames(m12)
```

# Combine data

```{r}
all_trips <- rbind(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)
```

#======================================================
# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
#======================================================

```{r}
colnames(all_trips)
```
```{r}
nrow(all_trips)
```
```{r}
dim(all_trips)
```
```{r}
head(all_trips)
```
```{r}
tail(all_trips)
```
```{r}
str(all_trips)
```
```{r}
summary(all_trips)
```

# Add columns that list the date, month, day, and year of each ride

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
# Add a "ride_length" calculation to all_trips 
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

# Inspect the structure of the columns
```{r}
str(all_trips)
```

# Convert "ride_length" from Factor to numeric
```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

# Remove "bad" data (ride length < 60 and docked bike)
```{r}
all_trips_v2 <- all_trips[!(all_trips$rideable_type == "docked_bike" |all_trips$ride_length<60),]
```

#=====================================
# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
#=====================================

# Descriptive analysis on ride_length
```{r}
summary(all_trips_v2$ride_length)
```
# Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
# See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
#Correct order for day of the week

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
# Run the code again
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

# analyze ridership data by type and weekday

```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()	 
  ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, number_of_rides)
```
# Visualize the number of rides by rider type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  geom_col(position = "dodge")

```

# Create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

#=================================================
# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
#=================================================

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'D:\\DH\\3rd - Semester II\\Google Data Analytics\\Case study\\Data\\CSV\\avg_ride_length.csv')
```
```{r}
ride_counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = length)
write.csv(ride_counts, file = 'D:\\DH\\3rd - Semester II\\Google Data Analytics\\Case study\\Data\\CSV\\number_of_rides.csv')
```


